---
title: "Fitness Tracker - A project in R"
output: html_notebook
---
## Business task
* Identify trends in smart device usage at competitor companies
* Apply trends in smart device usage to bellabeat customers
* Produce high level recommendations for bellabeat marketing strategy

## Data source
The project uses FitBit Fitness Tracker Dataset on Kaggle, made available through Mobius

#### Details:
* 30 FitBit users consented to tracking as part of distributed survey via Amazon Mechanical Turk for a month period in 2016
* The dataset includes 18 csv files containing: daily calories, steps, activity intensity and sleep; hourly calories, intensities and steps; minutely calories, intensities, METs, sleep and steps; user inputted weight data
* The dataset is open source and all participants consented to their data being used

#### Considerations
* Metadata from Kaggle was missing but has been located
* Sample size is 30 people over 1 month. This questions the validity of any generalisations about trends in the data
* Metadata from kaggle state 30 users, original study states 35, dataset contains 33
* There is no information about the 30 participants, meaning the data is likely to be biased

# Analysis

## Install packages
```{r echo=TRUE, message=FALSE, warning=FALSE}
install.packages("tidyverse")
library(tidyverse)
install.packages("janitor")
library(janitor)
```

## Read in data from csv
```{r echo=TRUE, message=FALSE, warning=FALSE}
daily_activity <- read_csv("C:/Users/katba/Data analysis learning/Portfolio/BellabeatCaseStudy/Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv")
daily_sleep <- read_csv("C:/Users/katba/Data analysis learning/Portfolio/BellabeatCaseStudy/Fitabase Data 4.12.16-5.12.16/sleepDay_merged.csv")
hourly_calories <- read_csv("C:/Users/katba/Data analysis learning/Portfolio/BellabeatCaseStudy/Fitabase Data 4.12.16-5.12.16/hourlyCalories_merged.csv")
hourly_intensity <- read_csv("C:/Users/katba/Data analysis learning/Portfolio/BellabeatCaseStudy/Fitabase Data 4.12.16-5.12.16/hourlyIntensities_merged.csv")
hourly_steps <- read_csv("C:/Users/katba/Data analysis learning/Portfolio/BellabeatCaseStudy/Fitabase Data 4.12.16-5.12.16/hourlySteps_merged.csv")
weight <- read_csv("C:/Users/katba/Data analysis learning/Portfolio/BellabeatCaseStudy/Fitabase Data 4.12.16-5.12.16/weightLogInfo_merged.csv")
```

## Explore daily dataframes 
```{r echo=TRUE, message=FALSE, warning=FALSE}
head(daily_activity)
head(daily_sleep)
str(daily_activity)
str(daily_sleep)
```

## Prepare daily dataframes to join
```{r echo=TRUE, message=FALSE, warning=FALSE}
daily_activity_clean <- daily_activity %>% 
  mutate(ActivityDate= as.Date(ActivityDate, format="%m/%d/%Y"))

daily_sleep_clean <- daily_sleep %>% 
  mutate(SleepDay= as.Date(SleepDay, format="%m/%d/%Y %H:%M:%S"))

daily_sleep_clean <- clean_names(daily_sleep_clean)
daily_activity_clean <- clean_names(daily_activity_clean)

daily_activity_clean <- rename(daily_activity_clean, date=activity_date)
daily_sleep_clean <- rename(daily_sleep_clean, date=sleep_day)
```

## Join daily dataframes
```{r echo=TRUE, message=FALSE, warning=FALSE}
daily <- full_join(daily_activity_clean, daily_sleep_clean, by = c("id", "date"))
```

## Additional join with weight 
```{r echo=TRUE, message=FALSE, warning=FALSE}
weight_clean <- clean_names(weight)
weight_clean <- weight_clean %>%
  mutate(date= as.Date(date, format="%m/%d/%Y %H:%M:%S"))
daily_all <- full_join(daily, weight_clean, by= c("id", "date"))
```

## Clean daily_all dataframe

### Review datatypes, min and max values
Datatypes, min and max values all make sense with the nature of the data
```{r echo=TRUE, message=FALSE, warning=FALSE}
summary(daily_all)
```

### Rename columns for consistency across dataset
```{r echo=TRUE, message=FALSE, warning=FALSE}
daily_all_clean <- rename(daily_all, total_distance_km=total_distance, tracker_distance_km=tracker_distance, 
                          logged_distance_km=logged_activities_distance, very_active_km=very_active_distance, 
                          moderate_active_km=moderately_active_distance, light_active_km=light_active_distance, 
                          sedentary_active_km=sedentary_active_distance, moderate_active_minutes=fairly_active_minutes,
                          total_minutes_in_bed=total_time_in_bed)
```

### Check for duplicated rows
Remove 3 duplicated rows
```{r echo=TRUE, message=FALSE, warning=FALSE}
sum(duplicated(daily_all_clean))
daily_all_clean <- daily_all_clean[!duplicated(daily_all_clean), ]
```
### Check number of unique ids for number of users
See 'considerations' for discussion
```{r echo=TRUE, message=FALSE, warning=FALSE}
length(unique(daily_all_clean$id))
```

### Check calculated columns
Total_distance is not tracker + logged, and it is also not just tracker for all rows. Unclear how total_distance was calculated, so tracker_distance will be used as primary measure. total_distance_km and total_distance_calc columns dropped.
```{r echo=TRUE, message=FALSE, warning=FALSE}
colnames(daily_all_clean)
daily_all_clean <- mutate(daily_all_clean, total_distance_calc=tracker_distance_km + logged_distance_km)
total_check <- ifelse(daily_all_clean$total_distance_km == daily_all_clean$total_distance_calc, "TRUE", "FALSE")
table(total_check)
daily_all_clean %>%
  filter(total_distance_km!=total_distance_calc) %>%
  arrange(logged_distance_km)
total_check2 <- ifelse(daily_all_clean$total_distance_km == daily_all_clean$tracker_distance_km, "TRUE", "FALSE")
table(total_check2)
```

```{r echo=TRUE, message=FALSE, warning=FALSE}
daily_all_clean <- subset(daily_all_clean, select= -c(total_distance_km,total_distance_calc))
```

### Check weight convertions
Check all weight measurements are accurate between kg and pound columns. 0.001 difference accounts for rounding differences in processing and will not impact future analysis.
```{r echo=TRUE, message=FALSE, warning=FALSE}
daily_all_clean <- mutate(daily_all_clean, weight_pounds_calc=weight_kg*2.20462)
daily_all_clean <- mutate(daily_all_clean, weight_pounds_dif= weight_pounds-weight_pounds_calc)
weight_check <- ifelse(daily_all_clean$weight_pounds_dif > 0.001, "TRUE", "FALSE")
table(weight_check)
```
## Clean hourly dataframes individually
```{r echo=TRUE, message=FALSE, warning=FALSE}
str(hourly_calories)
str(hourly_intensity)
str(hourly_steps)
```

### Convert characters to datetime
```{r echo=TRUE, message=FALSE, warning=FALSE}

hourly_steps$activity_hour <- as.POSIXct(hourly_steps$activity_hour, format="%m/%d/%Y %I:%M:%S %p")
hourly_calories$activity_hour <- as.POSIXct(hourly_calories$activity_hour, format="%m/%d/%Y %I:%M:%S %p")
hourly_intensity$activity_hour <- as.POSIXct(hourly_intensity$activity_hour, format="%m/%d/%Y %I:%M:%S %p")

```


### Clean column names
```{r echo=TRUE, message=FALSE, warning=FALSE}
library(janitor)
hourly_calories <- clean_names(hourly_calories)
hourly_intensity <- clean_names(hourly_intensity)
hourly_steps <- clean_names(hourly_steps)
```

### Check min/max values are accurate
```{r echo=TRUE, message=FALSE, warning=FALSE}
summary(hourly_calories)
summary(hourly_intensity)
summary(hourly_steps)
```
# Exploratory data analysis
## Investigating frequency of use by feature
Create dataframe with number of days of use per user for selected features
```{r echo=TRUE, message=FALSE, warning=FALSE}

daily_all_clean <- daily_all_clean %>% mutate_at(c('logged_distance_km'), ~na_if(., 0))

days_tracking_per_user <- daily_all_clean %>%
  group_by(id) %>%
  summarise(days_tracking_steps = sum(!is.na(total_steps)), days_logging_activity = sum(!is.na(logged_distance_km)), days_tracking_weight = sum(!is.na(weight_kg)), days_tracking_sleep = sum(!is.na(total_minutes_asleep)))

```

Plot average days of monthly usage
```{r echo=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
mean_days_tracking_per_user <- days_tracking_per_user %>%
  pivot_longer(everything(), names_to = "X", values_to = "Y") %>%
  group_by(X) %>%
  summarise(Mean = mean(Y, na.rm=TRUE)) 

mean_days_tracking_per_user <- mean_days_tracking_per_user[-5,]
mean_days_tracking_per_user <- mean_days_tracking_per_user %>%
  mutate(X = recode(X, "days_logging_activity" = "Manually logging activity", "days_tracking_sleep" = "Sleep tracking", "days_tracking_steps" = "Step tracking", "days_tracking_weight" = "Weight tracking")) %>%
  arrange(Mean)

mean_days_tracking_per_user %>%
  ggplot(aes(x=reorder(X, -Mean), y=Mean)) +
  geom_col(fill="skyblue") +
  labs(title="Average number of days feature is used over a month", x="Feature", y="Average number of days") +
  theme_classic()


```

## Invesigating each variable over the month
```{r echo=TRUE, message=FALSE, warning=FALSE}
daily_all_clean <- subset(daily_all_clean, select= -c(weight_pounds_calc, weight_pounds_dif))
```

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
col_names <- colnames(daily_all_clean)

for (i in col_names){
    plot <- ggplot(daily_all_clean, aes_string(x=daily_all_clean$date, y=i)) +
    geom_smooth()
    print(plot)
}

```
Observations from line graphs:\
* No clear pattern over the month for any variable\
* Slight downward trend towards the end of the testing period for some variables - however this is likely to be due to testing period ending\

## Investigating correlations between variables
Histogram shows at least one variable is not normally distributed, so spearman method used
```{r echo=TRUE, message=FALSE, warning=FALSE}
ggplot(daily_all_clean, aes(very_active_minutes)) +
  geom_histogram(binwidth=5)

```

```{r echo=TRUE, message=FALSE, warning=FALSE}
install.packages("Hmisc")
library(Hmisc)
install.packages("corrplot")
library(corrplot)
cor_daily <- cor(num_only_daily, use="pairwise.complete.obs", method="spearman")
corrplot(cor_daily, method="color", diag=FALSE, type="upper", tl.cex = 0.6)
```
Observations from correlations:\
* Fat column disregarded due to insufficient records, hence strong correlation with id\
* Expected correlations between steps, distance and activity levels\
* No further correlations to investigate at this point\
* Unexpected relationship between calories burnt and weight - this is likely due to limited data in weight fields\

## Investigating steps over time by hour
```{r echo=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
hourly_steps$time <- format(as.POSIXct(hourly_steps$activity_hour), format="%H:%M:%S")
av_steps_per_hour <- hourly_steps %>%
  group_by(time) %>%
  summarise(average_steps= mean(step_total))
```

```{r echo=TRUE, message=FALSE, warning=FALSE}
ggplot(data=av_steps_per_hour) +
  geom_point(mapping=aes(x=time, y=average_steps)) +
  theme_minimal() +
  labs(title= "Average number of steps through the day", x="Time of day", y="Average number of steps") +
  scale_x_discrete(guide=guide_axis(check.overlap=TRUE))


```
Observations:/
* People least active from 8pm to 11pm - targetted time for marketing

## Read in dataframes for HR and steps by minute
```{r echo=TRUE, message=FALSE, warning=FALSE}
hr <- read_csv("C:/Users/katba/Data analysis learning/Portfolio/BellabeatCaseStudy/Fitabase Data 4.12.16-5.12.16/heartrate_seconds_merged.csv")
steps <- read_csv("C:/Users/katba/Data analysis learning/Portfolio/BellabeatCaseStudy/Fitabase Data 4.12.16-5.12.16/minuteStepsNarrow_merged.csv")
```

## Clean HR and steps to prepare for join
```{r echo=TRUE, message=FALSE, warning=FALSE}
hr <- clean_names(hr)
steps <- clean_names(steps)

steps$activity_minute <- as.POSIXct(steps$activity_minute, format="%m/%d/%Y %I:%M:%S %p")
hr$time <- as.POSIXct(hr$time, format="%m/%d/%Y %I:%M:%S %p")

```

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
steps <- rename(steps, time = activity_minute)
```

## Change HR to by minute
Group by id, then average HR for each minute
```{r echo=TRUE, message=FALSE, warning=FALSE}
hr_min <- hr %>%
  group_by(id, time = cut(time, breaks= "1 min")) %>%
  summarise(value= mean(value))
```

Change time to datetime
```{r echo=TRUE, message=FALSE, warning=FALSE}

glimpse(hr_min)
hr_min$time <- as.character(hr_min$time)
hr_min$time <- as.POSIXct(hr_min$time, format="%Y-%m-%d %H:%M:%S")

```

Join HR and steps
```{r echo=TRUE, message=FALSE, warning=FALSE}
hr_steps <- inner_join(hr_min, steps, by = c("id", "time"))


```

## Check distribution of variables in preparation for correlation
```{r echo=TRUE, message=FALSE, warning=FALSE}
ggplot(hr_steps, aes(value)) +
  geom_histogram(binwidth=5)

ggplot(hr_steps, aes(steps)) +
  geom_histogram(binwidth=5)

```

```{r echo=TRUE, message=FALSE, warning=FALSE}
cor(hr_steps$value, hr_steps$steps, method="spearman")

```
```{r echo=TRUE, message=FALSE, warning=FALSE}
ggplot(hr_steps, aes(value, steps)) +
  geom_point(size=0.2)
```
Findings:/
* There is a weak correlation between steps and HR, this suggests fitbit users are using their watches for a range of exercise types. If there was a strong correlation we could assume fitbit users were mainly walking/running for exercise. This result shows some users will be wearing their watch during other types of exercise, such as gym activities.


Ideas for analysis:

